---
title: "Analyzing earthquake data using eqr"
author: "Slava Kohut"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Analyzing earthquake data using eqr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(eqr)
library(dplyr)
library(ggplot2)
data("eq_data")
```

## Intro

The _eqr_ package is a set of analytical tools designed for
working with the _NOAA Significant Earthquakes dataset_.
The workflow basically consists of the following steps:

- load and clean the data
- analyze the data using timeline plots and maps

In what follows, we will illustrate each of these steps by providing specific examples 
that feature _eqr_ functions. 

## Data cleaning

This step is performed using the `eq_clean_data`. The user needs to provide 
a path to the data set and the corresponding file name:
```{r, eval = FALSE}
eq_data <- eq_clean_data('./', 'noaa-data-set.txt')
```
As a result, one obtains a clean, ready-to-use data set that is compatible with plotting
and mapping functions. The `eq_clean_data` includes a call to `eq_location_clean` whose goal
is to extract a country name from the location string. The function is vectorized.
```{r}
eq_location_clean('CANADA: OTTAWA')
eq_location_clean(c('CANADA: OTTAWA', 'COLOMBIA: BOGOTA'))
```

## Mapping

Earthquake maps can be produced using the `eq_map` function.

```{r, fig.width = 7, fig.height = 5}
eq_data %>% filter(COUNTRY == 'COLOMBIA') %>% eq_map()
```

These maps are fully interactive and are annotated using earthquake dates by default.
Pointers' labels can be changed by supplying a column name that is used for generating a
new set of labels (the `annot_col` parameter). Let's use the earthquake magnitude for annotations:

```{r, fig.width = 7, fig.height = 5}
eq_data %>% filter(COUNTRY == 'COLOMBIA') %>% eq_map(annot_col = 'EQ_PRIMARY')
```

A more informative label (location + magnitude + the number of deaths) can be produced using 
the `eq_create_label` function:

```{r, fig.width = 7, fig.height = 5}
eq_data %>% filter(COUNTRY == 'COLOMBIA' & DATE > as.Date('1990-01-01')) %>%
  mutate(labels = eq_create_label(COUNTRY, EQ_PRIMARY, DEATHS)) %>%
  eq_map(annot_col = 'labels')
```

## Timeline plots

Timeline plots are a convenient way to visualize the earthquake sequential order along with
other important information for a particular country. These plots are generated using 
two _ggplot2_ geom functions: (i) `geom_timeline` and (ii) `geom_timeline_label`. 
As an example, let's generate a timeline plot for earthquakes happened in Italy.
The point size and color will be used for coding the earthquake magnitude and the number of deaths,
respectively.

```{r, fig.width = 7, fig.height = 5}
ggplot(data = eq_data %>% filter(COUNTRY == "ITALY"),
       aes(x = DATE, size = EQ_PRIMARY, color = TOTAL_DEATHS,
           xmin = as.Date('1950-01-01'),
           xmax = as.Date('2015-01-01'))) +
 geom_timeline() +
 theme_eq
```

Note that the data for more than one country can be plotted. Also, it is important to note that
the input data can be pre-filtered using the `xmin` and `xmax` parameters. The number of output points'
labels is controlled by the 'nmax' parameter which refers to the eathquake magnitude sorted in a descending
order.

```{r, fig.width = 7, fig.height = 5}
ggplot(data = eq_data %>% filter(COUNTRY %in% c('COLOMBIA', 'MEXICO', 'USA')),
       aes(x = DATE, y = COUNTRY, size = EQ_PRIMARY,
           color = TOTAL_DEATHS, xmin = as.Date('1970-01-01'),
           xmax = as.Date('2015-01-01'))) +
  geom_timeline_label(aes(label = as.character(DATE)), nmax = 2) +
  geom_timeline() +
  theme_eq
```
